Software variability in fMRI analysis:\\ comparing between-tool and numerical errors

### 2021-12-07 paper/paper.tex
Updates in introduction

Software variability broadly affects functional MRI analyses due to
differences among analytical tools and numerical instabilities. While
numerical and between-tool variability were both shown to impact analysis
outcomes, the extent to which these sources of variability compare for a
given analysis remains understudied. We extended a previous comparison
between fMRI analysis tools FSL, AFNI, and SPM, to measure numerical
variability through Monte-Carlo arithmetic. We found that in group analyses
between-tool variability was consistently larger than machine error,
whereas in subject analyses machine error approached between-tool
variability in some cases. Furthermore, between-tool variability and
machine error appeared moderately correlated in all tested conditions.
Finally, we found that numerical error at the precision of 17 bits --- a
precision level found in commonly-used libraries --- had effects of similar
magnitude \tristan{check that} than between-tool variability. Our findings motivate the
continued investigation of numerical instability in neuroimaging, and position it 
as a possible proxy for between-tool variations.
\end{abstract}

@@

Software variability has been highlighted as an important source of error
in fMRI analysis, originating from mainly two factors: Between-tool
variability denotes differences observed in results produced by different
software toolboxes --- such as FSL, AFNI and SPM --- when analyzing the same dataset
using similar methods. Numerical variability, in contrast, denotes
differences observed in results produced by the same analytical toolbox
executed in different environments – such as different operating systems or
different hardware platforms. This manuscript investigates the extent to
which both types of variability relate to each other.

Numerical variability stems from errors introduced in floating point
computations by numerical rounding, cancellation and absorption~\cite{muller2018handbook}. In
practice, the occurrence of such errors is affected by variations in
hardware, software dependencies, compilation, parallelization, and most
likely other factors. In general, the magnitude of such errors and their
variations is in the range of 1 unit in the last place (ulp) --- also known
as machine error, the difference between two successive floating-point
numbers. However, when computations depend on imprecise libraries,
numerical error may vary by several ulps. For instance, the numerical error
of the complementary error function (erfc) is larger than 3 ulps in several
mathematical libraries~\cite{zimmermann:hal-03141101}, implying that an update of this function could introduce
numerical perturbations larger than 3 ulps. Due to numerical instability,
machine error may amplify during its propagation through the analysis,
resulting in considerable loss of precision in final results. In fMRI,
numerical variability produced by Linux updates was shown to result in Dice
coefficients ranging from 0.0 to 1.0 for ICA components produced by FSL
MELODIC~\cite{Glatard2015}.

Between-tool variability is not a surprising concept for software
developers: different programs executed on identical data may obviously
produce different results. However, in fMRI, analysis programs all aim to
estimate brain activity from measured variations in the BOLD signal, which
is not expected to be measurably impacted by computational artifacts.
Therefore, the observation reported in~\cite{bowring2019exploring} that the three
main fMRI analysis toolboxes may result in marked
analytical differences, such as Dice coefficients ranging from 0.000 to
0.684 in comparisons of thresholded group-level statistic maps, has raised
substantial concern. In a follow-up study~\cite{bowring2021isolating}, the same
authors isolate the analytical components where the pipelines diverge,
highlighting a major impact of the signal model and suggesting that the
root-cause for between-tool variability depends on the analysis design and
task paradigm.  \tristan{[refer to Xinhui's pre-print].}

>>>>

Software variability broadly affects functional MRI analyses due to
differences among analytical tools as well as numerical instabilities. However,
the extent to which numerical and between-tool variabilities compare for a
given analysis is unclear. We extended a previous comparison
between fMRI analysis tools FSL, AFNI, and SPM, to measure numerical
variability through Monte-Carlo arithmetic. We found that, in group analyses,
between-tool variability was consistently larger than machine error,
whereas in subject analyses, machine error approached between-tool
variability in some cases. Furthermore, between-tool variability and
machine error appeared moderately correlated in all tested conditions.
Finally, we also found that numerical error at the precision of 17 bits --- a
precision level found in commonly-used libraries --- had effects of similar
magnitude \tristan{check that} than those of between-tool variability. Our findings motivate the
continued investigation of numerical instability in neuroimaging, and position it 
as a possible proxy for between-tool variations.
\end{abstract}

@@

Software variability has been highlighted as an important source of error
in fMRI analysis, originating from mainly two factors: between-tool
variability denotes differences in the outcomes of different
software toolboxes --- such as FSL, AFNI and SPM --- when analyzing the same dataset
using similar methods. Numerical variability, in contrast, denotes
differences in outcomes of the same analytical toolbox
executed in different environments – such as different operating systems or
different hardware platforms. This manuscript investigates the extent to
which both types of variability relate to each other.

Numerical variability stems from errors introduced in floating point
computations by numerical rounding, cancellation and absorption~\cite{muller2018handbook}. In
practice, the occurrence of such errors is affected by variations in
hardware, software dependencies, compilation, parallelization, and most
likely other factors. In general, the magnitude of such errors and their
variations is in the range of 1 unit in the last place (ulp) --- also known
as machine error, the difference between two successive floating-point
numbers. However, when computations depend on imprecise libraries,
numerical error may vary by several ulps. For instance, the numerical error
of the complementary error function (erfc) is larger than 3 ulps in several
mathematical libraries~\cite{zimmermann:hal-03141101}, implying that an update of this function could introduce
numerical perturbations larger than 3 ulps. Due to numerical instability,
machine error may amplify during its propagation through the analysis,
resulting in considerable loss of precision in final results. For instance, in fMRI,
numerical variability produced by Linux updates was shown to result in Dice
coefficients ranging from 0.0 to 1.0 for ICA components produced by FSL
MELODIC~\cite{Glatard2015}.

Between-tool variability is not a surprising concept: different programs
executed on the same data may obviously produce different results. However,
in fMRI analysis, programs all aim to estimate brain activity from measured
variations in the BOLD signal, which is not expected to be measurably
impacted by computational artifacts. Therefore, the observation reported
in~\cite{bowring2019exploring} that the three main fMRI analysis toolboxes
may result in marked analytical differences (Dice coefficients ranging from
0.000 to 0.684 for thresholded activation maps), has raised substantial
concern. In a follow-up study~\cite{bowring2021isolating}, the same authors
isolated the analytical components where the pipelines diverge,
highlighting a major impact of the signal model and suggesting that the
root-cause for between-tool variability depends on the analysis design and
task paradigm. Further, the work in~\cite{Li2021.12.01.470790} found
moderate agreement between results produced by five independent fMRI
analysis pipelines and analyzed in details the components that led to these
variations. Between-tool variability is supposedly an important determinant
of the substantial analytical variability recently observed in
fMRI~\cite{botvinik2020variability}.
